---
title: "DataViz for S/SE Asia"
author: 
  - name: Alex Diaz-Papkovich
affiliation:
  - &cruk Quantitative Life Sciences, McGill University,  Montreal, Canada
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output:
  html_notebook:
  df_print: paged
code_folding: show
toc: yes
toc_float: 
  collapsed: false
smooth_scroll: false
---
  
## Load libraries and set directories
```{r, message=FALSE, warning=FALSE}
library(cowplot)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(ggridges)
library(magrittr)
```

## Import data
```{r, message=FALSE, warning=FALSE}
data_path <- "/Volumes/Stockage/alex/ukbb_phenos/south_east_asia_phenos_2018328174511.csv"

# Import the main data
ukbb_data <- read.csv(data_path, stringsAsFactors = F)

# Reduce to just the countries we're interested in
pops <- c("Japan","Philippines","Malaysia")
ukbb_data_ese_asia <- ukbb_data[ukbb_data$COB %in% pops | ukbb_data$eth_txt=="Chinese",]

# Add a "populations" variable (since Chinese is an ethnicity, while the country of birth is technically not)
attach(ukbb_data_ese_asia)
ukbb_data_ese_asia$Population <- ifelse(eth_txt=="Chinese","Chinese",COB)
detach(ukbb_data_ese_asia)

ukbb_data_ese_asia_with_fev1 <- subset(ukbb_data_ese_asia, is.na(ukbb_data_ese_asia$X3063_0_0)==F)
```

# Get some summary stats
```{r}
table(ukbb_data_ese_asia_with_fev1[,c('Population',"sex")])

temp_malaysia <- ukbb_data_ese_asia_with_fev1[ukbb_data_ese_asia_with_fev1$Population=="Malaysia",]
temp_japan <- ukbb_data_ese_asia_with_fev1[ukbb_data_ese_asia_with_fev1$Population=="Japan",]
temp_philippines <- ukbb_data_ese_asia_with_fev1[ukbb_data_ese_asia_with_fev1$Population=="Philippines",]
```

# Regression using height, sex, age
```{r}
ukbb_data_ese_asia_with_fev1$fev1 <- ukbb_data_ese_asia_with_fev1$X3063_0_0
ukbb_data_ese_asia_with_fev1$sex_factor <- as.factor(ukbb_data_ese_asia_with_fev1$sex)
ukbb_data_ese_asia_with_fev1$pop_factor <- as.factor(ukbb_data_ese_asia_with_fev1$Population)

mod <- lm(data = ukbb_data_ese_asia_with_fev1, fev1 ~ Height + age + age^2 + sex_factor)#  + pop_factor)

summary(mod)

# Regress out sex, age, age^2, height
ukbb_data_ese_asia_with_fev1$fev_residuals <- mod$residuals
```

# Regression using height, age (no sex)
```{r}
mod <- lm(data = ukbb_data_ese_asia_with_fev1, fev1 ~ Height + age + age^2)

summary(mod)

# Regress out age, age^2, height
ukbb_data_ese_asia_with_fev1$fev_residuals_nosex <- mod$residuals
```

# Statistical testing (split by sex, regression includes sex)
```{r}
# Split into a list of data frames (one female, one male) so we can more easily run the tests
# We want graphics split by sex anyway
# Run Welch's t-tests (assuming unequal variances, different sample sizes)

# Select the dataset and variable of interest. Assumes sex is a binary chr ("0" for female, "1" for male)
testing_set <- ukbb_data_ese_asia_with_fev1
testing_var <- "fev_residuals"

testing_data <- list()
t_test_results <- list(list)

# First subset by sex, then remove missing data.
# There's probably a more succinct approach but this is readable and it works.
testing_data[["female"]] <- testing_set[testing_set$sex=="0",]
testing_data[["female"]] <- testing_data[["female"]][!is.na(testing_data[["female"]][[testing_var]]),]

testing_data[["male"]] <- testing_set[testing_set$sex=="1",]
testing_data[["male"]] <- testing_data[["male"]][!is.na(testing_data[["male"]][[testing_var]]),]

# Compare them relative to the Japanese population (we're not doing a huge pairwise thing)
# The Japanese cluster seems to be the only one with a different distribution visually
pop_vec = c("Chinese","Malaysia","Philippines")
sex_vec = c("female","male")

compare_pop = "Japan"

for (s in sex_vec){
 for (p in pop_vec){
   print(paste("Now comparing",p,"to",compare_pop))
   baseline_set <- testing_data[[s]][testing_data[[s]]$Population==compare_pop,]
   compare_set <- testing_data[[s]][testing_data[[s]]$Population==p,]
   
   t_test_results[[s]][[p]] <- t.test(baseline_set[[testing_var]], compare_set[[testing_var]])
 } 
}

```

# Statistical testing (both sexes, regression includes sex)
```{r}
# Run Welch's t-tests (assuming unequal variances, different sample sizes)

# Select the dataset and variable of interest
testing_set <- ukbb_data_ese_asia_with_fev1
testing_var <- "fev_residuals"

testing_data_both <- list()
t_test_results_both <- list()


testing_data_both[["both"]] <- testing_set
testing_data_both[["both"]] <- testing_data_both[["both"]][!is.na(testing_data_both[["both"]][[testing_var]]),]

# Compare them relative to the Japanese population (we're not doing a huge pairwise thing)
# The Japanese cluster seems to be the only one with a different distribution visually
pop_vec = c("Chinese","Malaysia","Philippines")
sex_vec <- c("both")

compare_pop = "Japan"
for (s in sex_vec){
  for (p in pop_vec){
   print(paste("Now comparing",p,"to",compare_pop))
   baseline_set <- testing_data_both[[s]][testing_data_both[[s]]$Population==compare_pop,]
   compare_set <- testing_data_both[[s]][testing_data_both[[s]]$Population==p,]
   
   t_test_results_both[[p]] <- t.test(baseline_set[[testing_var]], compare_set[[testing_var]])
  } 
}

```

# Kind of a pain but we need to merge the various statistics
```{r}
t_test_df <- data.frame(sex=character(),
                        Population=character(),
                        mean_diff=double(),
                        pval=double(),
                        ast=character(),
                        stringsAsFactors = F)
sex_vec <- c("female","male")

for (s in sex_vec){
  for (p in pop_vec){
    # Difference of means
    mean1 <- t_test_results[[s]][[p]]$estimate[1]
    mean2 <- t_test_results[[s]][[p]]$estimate[2]
    
    mean_diff <- mean1 - mean2
    
    # Find the p-value and adjust it for multiple testing
    pval <- t_test_results[[s]][[p]]$p.value
  
    if (pval < 0.05/length(pop_vec)){
      ast <- "*"
    } else {
      ast <- ""
    }
    
    # Initialize the data frame (first observation)
    if(p==pop_vec[1] & s==sex_vec[1]){
        t_test_df[1,] <- list(as.character(s),
                              as.character(p),
                              as.numeric(mean_diff), 
                              as.numeric(pval),
                              as.character(ast))
    } else {
      t_test_df <- rbind(t_test_df, list(as.character(s),
                                         as.character(p),
                                       as.numeric(mean_diff), 
                                       as.numeric(pval),
                                       as.character(ast)))
    }
  }
}
# Insert the Japanese row and sort alphabetically
t_test_df <- rbind(t_test_df, list(as.character("female"),
                                   as.character("Japan"),
                                   as.numeric(0),
                                   as.numeric(0),
                                   as.character("")))
t_test_df <- rbind(t_test_df, list(as.character("male"),
                                   as.character("Japan"),
                                   as.numeric(0),
                                   as.numeric(0),
                                   as.character("")))

# Round values
#t_test_df$pval <- round(t_test_df$pval, 2)

# Sort to match labelling
t_test_df <- t_test_df[order(t_test_df$Population),]
t_test_df[order(t_test_df$Population),]
```

# Ridge plot of females
```{r}
# Subset to females
plotting_data <- ukbb_data_ese_asia_with_fev1[ukbb_data_ese_asia_with_fev1$sex=="0",]

# Remove NaN
plotting_data <- plotting_data[!is.na(plotting_data$fev_residuals),]

# Get means (sort by these)
plotting_means <- aggregate(plotting_data$fev_residuals, list(plotting_data$Population), mean)
names(plotting_means) <- c("Population", "mean")

# Colour palette
pal <- c("#e377c2","#98df8a","#aec7e8","#8c564b")

# Subset t-test results
t_test_vals <- t_test_df[t_test_df$sex=="female",]

label_vec <- paste(names(table(plotting_data$Population)),
                                   " (n=",
                                   table(plotting_data$Population),
                                   ", μ=", formatC(plotting_means$mean, digits=2),
                                   ", p-val=", formatC(t_test_vals$pval, format="e", digits=2),
                                   t_test_vals$ast, ")",
                                   sep="")
# cheating a bit here
label_vec[2] <- paste("Japan (n=202, μ=",formatC(plotting_means[plotting_means$Population=="Japan",]$mean, digits=2),")",sep="")

p_female <- ggplot(plotting_data, aes(x=fev_residuals, y=Population,fill=Population)) +
  geom_density_ridges(scale=2, alpha=0.4) +
  scale_fill_manual(values=pal,
                    labels = label_vec
                    ) +
  geom_vline(xintercept=plotting_means$mean, linetype="dashed", colour=pal, size=1) +
  xlim(-3.5,3.5) +
  ggtitle("Residual FEV1 (female)") +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        legend.text.align = 0) +
  guides(fill = guide_legend(reverse=TRUE))

ggsave("/Volumes/Stockage/alex/ukbb_images/other/fev1_residual_east_asia_female_ridges.png", p_female, width=15)
p_female
```

# Ridge plot of males
```{r}
# Subset to males
plotting_data <- ukbb_data_ese_asia_with_fev1[ukbb_data_ese_asia_with_fev1$sex=="1",]

# Remove NaN
plotting_data <- plotting_data[!is.na(plotting_data$fev_residuals),]

# Get means (sort by these)
plotting_means <- aggregate(plotting_data$fev_residuals, list(plotting_data$Population), mean)
names(plotting_means) <- c("Population", "mean")

# Colour palette
pal <- c("#e377c2","#98df8a","#aec7e8","#8c564b")

# Subset t-test results
t_test_vals <- t_test_df[t_test_df$sex=="male",]

label_vec <- paste(names(table(plotting_data$Population)),
                                   " (n=",
                                   table(plotting_data$Population),
                                   ", μ=", formatC(plotting_means$mean, digits=2),
                                   ", p-val=", formatC(t_test_vals$pval, format="e", digits=2),
                                   t_test_vals$ast, ")",
                                   sep="")

label_vec[2] <- paste("Japan (n=30, μ=",formatC(plotting_means[plotting_means$Population=="Japan",]$mean, digits=2),")",sep="")

p_male <- ggplot(plotting_data, aes(x=fev_residuals, y=Population,fill=Population)) +
  geom_density_ridges(scale=2, alpha=0.4) +
  scale_fill_manual(values=pal,
                    labels = label_vec
                    ) +
  geom_vline(xintercept=plotting_means$mean, linetype="dashed", colour=pal, size=1) +
  xlim(-3.5,3.5) +
  ggtitle("Residual FEV1 (male)") +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank()) +
  guides(fill = guide_legend(reverse=TRUE))

ggsave("/Volumes/Stockage/alex/ukbb_images/other/fev1_residual_east_asia_male_ridges.png", p_male, width=15)
p_male
```

# Put the two plots on the same graphic (cowplot)
```{r}
p_combined <- plot_grid(p_female, p_male, align="v", nrow=2, ncol=1)
p_combined
ggsave("/Volumes/Stockage/alex/ukbb_images/other/fev1_residual_east_asia_ridges.png", p_combined,
       height=7, width=10)
```

# Same but use residuals from regression that didn't include sex
```{r}
# Split into a list of data frames (one female, one male) so we can more easily run the tests
# We want graphics split by sex anyway
# Run Welch's t-tests (assuming unequal variances, different sample sizes)

# Select the dataset and variable of interest. Assumes sex is a binary chr ("0" for female, "1" for male)
testing_set <- ukbb_data_ese_asia_with_fev1
testing_var <- "fev_residuals_nosex"

testing_data <- list()
t_test_results <- list(list)

# First subset by sex, then remove missing data.
# There's probably a more succinct approach but this is readable and it works.
testing_data[["female"]] <- testing_set[testing_set$sex=="0",]
testing_data[["female"]] <- testing_data[["female"]][!is.na(testing_data[["female"]][[testing_var]]),]

testing_data[["male"]] <- testing_set[testing_set$sex=="1",]
testing_data[["male"]] <- testing_data[["male"]][!is.na(testing_data[["male"]][[testing_var]]),]

# Compare them relative to the Japanese population (we're not doing a huge pairwise thing)
# The Japanese cluster seems to be the only one with a different distribution visually
pop_vec = c("Chinese","Malaysia","Philippines")
sex_vec = c("female","male")

compare_pop = "Japan"

for (s in sex_vec){
 for (p in pop_vec){
   print(paste("Now comparing",p,"to",compare_pop))
   baseline_set <- testing_data[[s]][testing_data[[s]]$Population==compare_pop,]
   compare_set <- testing_data[[s]][testing_data[[s]]$Population==p,]
   
   t_test_results[[s]][[p]] <- t.test(baseline_set[[testing_var]], compare_set[[testing_var]])
 } 
}

```

# Merge the various statistics
```{r}
t_test_df <- data.frame(sex=character(),
                        Population=character(),
                        mean_diff=double(),
                        pval=double(),
                        ast=character(),
                        stringsAsFactors = F)
for (s in sex_vec){
  for (p in pop_vec){
    # Difference of means
    mean1 <- t_test_results[[s]][[p]]$estimate[1]
    mean2 <- t_test_results[[s]][[p]]$estimate[2]
    
    mean_diff <- mean1 - mean2
    
    # Find the p-value and adjust it for multiple testing
    pval <- t_test_results[[s]][[p]]$p.value
  
    if (pval < 0.05/length(pop_vec)){
      ast <- "*"
    } else {
      ast <- ""
    }
    
    # Initialize the data frame (first observation)
    if(p==pop_vec[1] & s==sex_vec[1]){
        t_test_df[1,] <- list(as.character(s),
                              as.character(p),
                              as.numeric(mean_diff), 
                              as.numeric(pval),
                              as.character(ast))
    } else {
      t_test_df <- rbind(t_test_df, list(as.character(s),
                                         as.character(p),
                                       as.numeric(mean_diff), 
                                       as.numeric(pval),
                                       as.character(ast)))
    }
  }
}
# Insert the Japanese row and sort alphabetically
t_test_df <- rbind(t_test_df, list(as.character("female"),
                                   as.character("Japan"),
                                   as.numeric(0),
                                   as.numeric(0),
                                   as.character("")))
t_test_df <- rbind(t_test_df, list(as.character("male"),
                                   as.character("Japan"),
                                   as.numeric(0),
                                   as.numeric(0),
                                   as.character("")))

# Round values
#t_test_df$pval <- round(t_test_df$pval, 2)

# Sort to match labelling
t_test_df <- t_test_df[order(t_test_df$Population),]
t_test_df[order(t_test_df$Population),]
```

# Ridge plot of females
```{r}
# Subset to females
plotting_data <- ukbb_data_ese_asia_with_fev1[ukbb_data_ese_asia_with_fev1$sex=="0",]

# Remove NaN
plotting_data <- plotting_data[!is.na(plotting_data$fev_residuals_nosex),]

# Get means (sort by these)
plotting_means <- aggregate(plotting_data$fev_residuals_nosex, list(plotting_data$Population), mean)
names(plotting_means) <- c("Population", "mean")

# Colour palette
pal <- c("#e377c2","#98df8a","#aec7e8","#8c564b")

# Subset t-test results
t_test_vals <- t_test_df[t_test_df$sex=="female",]

label_vec <- paste(names(table(plotting_data$Population)),
                                   " (n=",
                                   table(plotting_data$Population),
                                   ", μ=", formatC(plotting_means$mean, digits=2),
                                   ", p-val=", formatC(t_test_vals$pval, format="e", digits=2),
                                   t_test_vals$ast, ")",
                                   sep="")
# cheating a bit here
label_vec[2] <- paste("Japan (n=202, μ=",formatC(plotting_means[plotting_means$Population=="Japan",]$mean, digits=2),")",sep="")

p_female <- ggplot(plotting_data, aes(x=fev_residuals_nosex, y=Population,fill=Population)) +
  geom_density_ridges(scale=2, alpha=0.4) +
  scale_fill_manual(values=pal,
                    labels = label_vec
                    ) +
  geom_vline(xintercept=plotting_means$mean, linetype="dashed", colour=pal, size=1) +
  #xlab("FEV1 residuals (female)") + ylab("Population") +
  xlim(-3.5,3.5) +
  ggtitle("Residual FEV1 (female)") +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        legend.text.align = 0) +
  guides(fill = guide_legend(reverse=TRUE))

ggsave("/Volumes/Stockage/alex/ukbb_images/other/fev1_residual_east_asia_female_ridges_nosex.png", p_female, width=15)
p_female
```

# Ridge plot of males
```{r}
# Subset to males
plotting_data <- ukbb_data_ese_asia_with_fev1[ukbb_data_ese_asia_with_fev1$sex=="1",]

# Remove NaN
plotting_data <- plotting_data[!is.na(plotting_data$fev_residuals_nosex),]

# Get means (sort by these)
plotting_means <- aggregate(plotting_data$fev_residuals_nosex, list(plotting_data$Population), mean)
names(plotting_means) <- c("Population", "mean")

# Colour palette
pal <- c("#e377c2","#98df8a","#aec7e8","#8c564b")

# Subset t-test results
t_test_vals <- t_test_df[t_test_df$sex=="male",]

label_vec <- paste(names(table(plotting_data$Population)),
                                   " (n=",
                                   table(plotting_data$Population),
                                   ", p-val=", formatC(t_test_vals$pval, format="e", digits=2),
                                   t_test_vals$ast, ")",
                                   sep="")

label_vec[2] <- paste("Japan (n=30, μ=",formatC(plotting_means[plotting_means$Population=="Japan",]$mean, digits=2),")",sep="")

p_male <- ggplot(plotting_data, aes(x=fev_residuals_nosex, y=Population,fill=Population)) +
  geom_density_ridges(scale=2, alpha=0.4) +
  scale_fill_manual(values=pal,
                    labels = label_vec
                    ) +
  geom_vline(xintercept=plotting_means$mean, linetype="dashed", colour=pal, size=1) +
  #xlab("FEV1 residuals (male)") + ylab("Population") +
  xlim(-3.5,3.5) +
  ggtitle("Residual FEV1 (male)") +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank()) +
  guides(fill = guide_legend(reverse=TRUE))

ggsave("/Volumes/Stockage/alex/ukbb_images/other/fev1_residual_east_asia_male_ridges_nosex.png", p_male, width=15)
p_male
```

# Put the two plots on the same graphic (cowplot)
```{r}
p_combined <- plot_grid(p_female, p_male, align="v", nrow=2, ncol=1)
p_combined
ggsave("/Volumes/Stockage/alex/ukbb_images/other/fev1_residual_east_asia_ridges_nosex.png", p_combined,
       height=7, width=10)
```
# Simple test
```{r}
# Run a t-test for the manuscript - we just need to know the effect size controlling for age, sex, age^2, height.
temp_set <- ukbb_data_ese_asia_with_fev1[!is.na(ukbb_data_ese_asia_with_fev1$fev_residuals),]
temp_set <- temp_set[temp_set$sex=="1",]

baseline_set <- temp_set[temp_set$Population=="Japan",]
compare_set <- temp_set[temp_set$Population=="Chinese",]

test_results <- t.test(baseline_set[["fev_residuals"]], compare_set[["fev_residuals"]])
test_results
test_results$estimate[1] - test_results$estimate[2]
```